Interpretable Neural Networks using EAGGA

### 2025-03-12 paper/main.tex
bullets on NAS

(evolutionary algorithms)
- another model-free class of optimisation algorithms, inspired by naturally occuring evolution
- based on a population, which iteratively (iteration = generation) generates $\lambda$ offspring via 3 operators (all sub-bullets from \citet[pp. 10-]{genetic_algos})
  * reproduction: pick an individual to reproduce with p proportional to its fitness
  * crossover
    - select 2 ``parents'' from pool of reproducing individuals
    - for a position $k$ in their HP configs, with certain p, swap their values after $k$, yielding 2 children
    %use example from book page 12
  * mutation: with certain p, change values of an individual, e.g. by adding Gaussian noise (Gaussian mutation) to real values, or flipping bit for binary values
- selection at end of each genertion: keep $\mu$ best individuals (based on some fitness function), either from only offspring (``$(\mu,\lambda)$-selection'')
  or (more commonly) from population + offspring (``$(\mu+\lambda)$-selection'', guaranteed to keep best individual)
- pro: conceptually simple + can handle even complex parameter spaces (continuous, discrete, hierarchical, etc.) given appropriate implementation of operators
all above starting at HPO from \citep[chap. 1.3]{feurer_hyperparameter_2019}
- prominent examples
  * CMA-ES
    - ``Covariance Matrix Adapation Evolution Strategy''
    - offspring generation exclusively via multivariate normal \citep[p. 8]{hansen2023cmaevolutionstrategytutorial}
      * mean = weighted average of previous generation
      * covariance = weighted covariance of previous generation, with weighting as for mean
    - weighing scheme done to sample in a way as to reproduce previously successful (i.e. selected) steps \citep[p. 11]{hansen2023cmaevolutionstrategytutorial}
  * differential evolution \citep[-]{differential_evolution}
    - init population randomly so that entire param space is covered
      * $D$-dim vectors $\boldsymbol{x}_{i,G}$ with $i=1,2,...,n$ for gen $G$
    - mutation
      * for each $\boldsymbol{x}_{i,G}$, generate \textbf{mutant vector} $\boldsymbol{v}_{i,G+1}=\boldsymbol{x}_{r_1,G}+F\cdot(\boldsymbol{x}_{r_2,G}-\boldsymbol{x}_{r_3,G})$
      * $r_1,r_2,r_3\in\{1,2,...,n\}$ mutually different random idx + different from $i$
      * constant $F\in[0,2]$
    - crossover
      * \textbf{trial vector} $\boldsymbol{u}_{i,G+1}=(u_{i,G+1}^{(1)},u_{i,G+1}^{(2)},...,u_{i,G+1}^{(D)})$
        - chose random index $R\in\{1,2,...,D\}$
        - sample $p\sim U[0,1]$
        - define crossover constant $\text{CR}\in[0,1]$
        - $u_{i,G+1} = \begin{dcases}
          v_{i,G+1}^{(j)} & \text{if } p\le\text{CR}\text{ or } j=R \\
          x_{i,G}^{(j)} & \text{else}
         \end{dcases}$
    - selection: greedy, compare fitness of $\boldsymbol{u}_{i,G+1}$ with $\boldsymbol{x}_{i,G}$, pick better

[model based]
- contrast to model free, fit a surrogate model on blackbox problem + optimise this
(Bayesian optimisation)
- iterative algo comprising of
  * surrogate model for black box problem, needs to be able to model mean + variance
  * acquisition function, to decide which point to query next
- in each iteration
  * fit surrogate model on all data points (posterior distribution)
  * get highest utility point from acquisition function for newly fitted surrogate model -> add to data points
  -> hence Bayesian, get new query point given already previously fitted points
  * acquisition function trades off exploration + exploitation of surrogate
  * BO up to here from \citep[chap. 1.3.2]{feurer_hyperparameter_2019} + \citep[pp. 2-3]{frazier2018tutorialbayesianoptimization}
- no optimising of model directly, instead iteratively optimise acquisition function
- common choices for surrogate models
  * GPs
    - pros
      * fully specified by mean + covariance
        - covariance function (aka kernel) solely determines quality of GP
        - kernel as function of two points from search space, yields their covariance
        - usual property of kernels: the closer points are in search space, the more strongly their correlation \citep[p. 5]{frazier2018tutorialbayesianoptimization}
      * well-calibrated uncertainty estimates
      * closed-form computability
    - con: neither scales well in \# of data points nor in \# of HP dimensions BUT workaround: sparse GPs, approx full GP with small subset of original dataset
  * random forests
    - can handle complex search spaces (high-dim, categorical, hierarchical), unlike GPs
    - computational complexity scales far better than GPs
      * GPs $O(n^3)$ fitting, $O(n^2)$ predicting
      * RFs $O(n \log n)$ fitting, $O(\log n)$ predicting
- BO stuff from \citep[chap. 1.3.2]{feurer_hyperparameter_2019}

\subsection{Neural Architecture Search (NAS)}
give background on NAS (e.g. one-shot, etc.), but for this project wanted to have a "minimal viable product" to test whether NNs can yield competitive performance
on tabular ML + be interpretable using EAGGA at all before optimising the inner networks to the max

>>>>

(evolutionary algorithms)
- another model-free class of optimisation algorithms, inspired by naturally occuring evolution
- based on a population, which iteratively (iteration = generation) generates $\lambda$ offspring via 3 operators (all sub-bullets from \citet[pp. 10-]{genetic_algos})
  * \textbf{reproduction}: pick an individual to reproduce with p proportional to its fitness
  * \textbf{crossover}
    - select 2 ``parents'' from pool of reproducing individuals
    - for a position $k$ in their HP configs, with certain p, swap their values after $k$, yielding 2 children
    %use example from book page 12
  * \textbf{mutation}: with certain p, change values of an individual, e.g. by adding Gaussian noise (Gaussian mutation) to real values, or flipping bit for binary values
- selection at end of each genertion: keep $\mu$ best individuals (based on some fitness function), either from only offspring (``$(\mu,\lambda)$-selection'')
  or (more commonly) from population + offspring (``$(\mu+\lambda)$-selection'', guaranteed to keep best individual)
- pro: conceptually simple + can handle even complex parameter spaces (continuous, discrete, hierarchical, etc.) given appropriate implementation of operators
all above starting at HPO from \citep[chap. 1.3]{feurer_hyperparameter_2019}
- prominent examples
  * CMA-ES
    - ``Covariance Matrix Adapation Evolution Strategy''
    - offspring generation exclusively via multivariate normal \citep[p. 8]{hansen2023cmaevolutionstrategytutorial}
      * mean = weighted average of previous generation
      * covariance = weighted covariance of previous generation, with weighting as for mean
    - weighing scheme done to sample in a way as to reproduce previously successful (i.e. selected) steps \citep[p. 11]{hansen2023cmaevolutionstrategytutorial}
  * differential evolution \citep[-]{differential_evolution}
    - init population randomly so that entire param space is covered
      * $D$-dim vectors $\boldsymbol{x}_{i,G}$ with $i=1,2,...,n$ for gen $G$
    - mutation
      * for each $\boldsymbol{x}_{i,G}$, generate \textbf{mutant vector} $\boldsymbol{v}_{i,G+1}=\boldsymbol{x}_{r_1,G}+F\cdot(\boldsymbol{x}_{r_2,G}-\boldsymbol{x}_{r_3,G})$
      * $r_1,r_2,r_3\in\{1,2,...,n\}$ mutually different random idx + different from $i$
      * constant $F\in[0,2]$
    - crossover
      * \textbf{trial vector} $\boldsymbol{u}_{i,G+1}=(u_{i,G+1}^{(1)},u_{i,G+1}^{(2)},...,u_{i,G+1}^{(D)})$
        - chose random index $R\in\{1,2,...,D\}$
        - sample $p\sim U[0,1]$
        - define crossover constant $\text{CR}\in[0,1]$
        - $u_{i,G+1} = \begin{dcases}
          v_{i,G+1}^{(j)} & \text{if } p\le\text{CR}\text{ or } j=R \\
          x_{i,G}^{(j)} & \text{else}
         \end{dcases}$
    - selection: greedy, compare fitness of $\boldsymbol{u}_{i,G+1}$ with $\boldsymbol{x}_{i,G}$, pick better

[model based]
- contrast to model free, fit a surrogate model on blackbox problem + optimise this
(Bayesian optimisation)
- iterative algo comprising of
  * surrogate model for black box problem, needs to be able to model mean + variance
  * acquisition function, to decide which point to query next
- in each iteration
  * fit surrogate model on all data points (posterior distribution)
  * get highest utility point from acquisition function for newly fitted surrogate model -> add to data points
  -> hence Bayesian, get new query point given already previously fitted points
  * acquisition function trades off exploration + exploitation of surrogate
  * BO up to here from \citep[chap. 1.3.2]{feurer_hyperparameter_2019} + \citep[pp. 2-3]{frazier2018tutorialbayesianoptimization}
- no optimising of model directly, instead iteratively optimise acquisition function
- common choices for surrogate models
  * \textbf{gaussian processes}
    - pros
      * fully specified by mean + covariance
        - covariance function (aka kernel) solely determines quality of GP
        - kernel as function of two points from search space, yields their covariance
        % TODO: write down kernel equation, cf. p 10ff https://ieeexplore.ieee.org/document/7352306
        - usual property of kernels: the closer points are in search space, the more strongly their correlation \citep[p. 5]{frazier2018tutorialbayesianoptimization}
      * well-calibrated uncertainty estimates
      * closed-form computability
    - con: neither scales well in \# of data points nor in \# of HP dimensions BUT workaround: sparse GPs, approx full GP with small subset of original dataset
  * \textbf{random forests}
    - can handle complex search spaces (high-dim, categorical, hierarchical), unlike GPs
    - computational complexity scales far better than GPs
      * GPs $O(n^3)$ fitting, $O(n^2)$ predicting
      * RFs $O(n \log n)$ fitting, $O(\log n)$ predicting
- common choices for acquisition functions
  % TODO: cf. p. 13 cf. p 10ff https://ieeexplore.ieee.org/document/7352306
  * \textbf{expected improvement}
  * \textbf{Thompson sampling}
- all BO stuff if not specified differently comes from \citep[chap. 1.3.2]{feurer_hyperparameter_2019}

\subsection{Neural Architecture Search (NAS)}
- approaches specifically for NN HPO due to flexible structure of NNs and thus very large search space (each layer can have different \# nodes, different activations, for CNN different pooling, etc. operations)
- we don't employ NAS for our extension as research focusses on the NLP and image domains, % TODO: find some more references
  whose datasets exhibit strong correlation amongst features (i.e. tokens or pixels, respectively), an effect that is much weaker for tabular data \citep[p. 1]{Borisov_2024}
- at least want to give a little overview over most notable approaches
  * \textbf{cell search space}
    - modularise NN architecture into cells, NN as chain-structure of those
    - cell = basic building block, fixed component of an NN
      * e.g. linear layer with specific \# of neurons in feedforward
      * or convolutional / pooling / etc layer for CNN
    - then optimise sequential placement of the cells as HPO problem
      * e.g. via random search or BO
      * \citet[p. 3]{zoph2017neuralarchitecturesearchreinforcement} + \citet[pp. 2-4]{Zoph_2018_CVPR} use RNN to recursively optimise HPs of a cell given already determined
        previous cells (+ HPs of current cell), which is trained via RL with the different HPs the RNN predicts being the actions and performance on held-out data being reward
    - \citep[chap. 3.2]{elsken_neural_2019}
  * \textbf{one-shot model}
    - trains one overall ``fabric'' comprising all architectures of search space
    - visualisation as DAG, nodes are nodes, edges in-between are operations on a node
    - each path through DAG (from input to output node) represents one architecture of search space -> train entire DAG, then pick optimal path (architecture)
    - pro: very efficient training, individual architectures share operations along edges they share, training one-shot model trains all subsumed models (more expensive
      than training a single model but less expensive than trying out all configurations included in the fabric)
    - \citep[pp. 1-2, p.8]{saxena2017convolutionalneuralfabrics}
    % don't necessarily talk about co-adaptation problem
